---
title: "Regression Equation Comparison"
description: |
  Seawater Analysis
output: 
  distill::distill_article:
    code_folding: hide
---

## Analysis of Seawater

### Overview
This analysis will look at the relationship between oxygen saturation of seawater off California’s coast and several physical and chemical variables including oxygen saturation, water temperature, water salinity, water depth, phosphate concentration, and nitrite concentrations.

Data citation: CalCOFI data are available for use without restriction. Data downloaded from https://calcofi.org/ccdata.html.  Accessed 1/10/2022.

```{r setup, include = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(AICcmodavg)
library(equatiomatic)
```

### Read in the data
```{r}
seawater_samples <- read_csv(here('data', 'calcofi_seawater_samples.csv'))
```

### Clean the data
```{r}
seawater_clean <- seawater_samples %>% 
  drop_na() %>% 
  rename(temp = t_deg_c, # variables are renamed so they are easier to work with and understand
         depth = depth_m,
         phosphate = po4u_m,
         nitrite = no2u_m)
```

### Create linear regression models using functions to store the equations
Model 1 is oxygen saturation as a function of water temperature, salinity, and phosphate concentration.

Model 2 is oxygen saturation as a function of water temperature, salinity, phosphate concentration, and depth.

A third model has also been added. Model 3 is oxygen saturation as a function of water temperature, salinity, phosphate concentration, nitrite concentration, and depth.

```{r}
f1 <- o2sat ~ temp + salinity + phosphate
mdl1 <- lm(f1, data = seawater_clean)

f2 <- o2sat ~ temp + salinity + phosphate + depth
mdl2 <- lm(f2, data = seawater_clean)

f3 <- o2sat ~ temp + salinity + phosphate + + nitrite + depth
mdl3 <- lm(f3, data = seawater_clean)
```

### Use AIC to compare the models
AIC = Akaike Information Criterion

Model with the lowest AIC is the preferred model. Ideally, the best model is lower than the next best model by at least 2.0. A difference of 2 indicates a significant difference in model fit.

```{r}
AICc(mdl1) # AICc() corrects for sample size

AICc(mdl2)

AICc(mdl3)

AICcmodavg::aictab(list(mdl1, mdl2, mdl3))
```

- AIC of model 1 = 619.0251

- AIC of model 2 = 616.6048

- AIC of model 3 = 613.5962

- The AIC of model 1 > model 2 > model 3.

- Model 3 is lower than model 2 by 3.01 and lower than model 1 by 5.43, so since it is lower by at least 2, model 3 is the preferred model.

- Model 3 is the preferred model via the AIC method.


### Use 10-fold cross validation to compare the models
Root-mean-square error is the scoring method.

```{r}
folds <- 10 # number of folds
fold_vec <- rep(1:folds, length.out = nrow(seawater_clean)) # fold vector is repeating over each fold
table(fold_vec)

set.seed(42) # allows others to get same set of random numbers if they try to replicate this work

seawater_fold <- seawater_clean %>% 
  mutate(group = sample(fold_vec, size = n(), replace = FALSE))
```

Create root-mean-square error function.

```{r}
calc_rmse <- function(x, y) {
  rmse_result <- (x-y)^2 %>% mean() %>% sqrt()
  return(rmse_result)
}
```

Calculate over all folds and take the average.

```{r}
rmse_df <- data.frame() # Create a blank data frame.

# Below is a for loop - loops through all 10 folds.
for(i in 1:folds) {
  kfold_test_df <-seawater_fold %>% 
    filter(group == i) # in group i
  kfold_train_df <-seawater_fold %>% 
    filter(group != i) # not in group i
  
  kfold_mdl1 <- lm(f1, data = kfold_train_df)
  kfold_mdl2 <- lm(f2, data = kfold_train_df)
  kfold_mdl3 <- lm(f3, data = kfold_train_df)
  
  kfold_pred_df <- kfold_test_df %>% 
    mutate(mdl1 = predict(kfold_mdl1, kfold_test_df),
           mdl2 = predict(kfold_mdl2, .), # the period is a shortcut that says you are predicting on the data frame
           mdl3 = predict(kfold_mdl3, .)) 
  kfold_rmse <- kfold_pred_df %>% 
    summarize(rmse_mdl1 = calc_rmse(mdl1, o2sat),
              rmse_mdl2 = calc_rmse(mdl2, o2sat),
              rmse_mdl3 = calc_rmse(mdl3, o2sat))
  
  #Store the last chunk above by combining with the blank data frame from above.
  rmse_df <- bind_rows(rmse_df, kfold_rmse)
}

rmse_df %>% 
  summarize(mean_rmse_mdl1 = mean(rmse_mdl1),
            mean_rmse_mdl2 = mean(rmse_mdl2),
            mean_rmse_mdl3 = mean(rmse_mdl3))

```

- Model 1 RMSE = 4.976605

- Model 2 RMSE = 4.876322

- Model 3 RMSE = 4.795063

- Model 3 has the lowest root-mean-square error meaning it has the lowest error when predicting data.

- Model 3 is the preferred model via the k-fold cross validation method using RMSE scoring.


**Both AIC and k-fold cross validation using root-mean-square error as the scoring method indicate model 3 to be the preferred model.**

### The coefficients for the final predictive model are identified:

```{r}
final_mdl <- lm(f3, data = seawater_clean)
summary(final_mdl)
```

### The final model:
`r equatiomatic::extract_eq(final_mdl, wrap = TRUE)`


Where:

- o2sat = oxygen saturation

- temp = water temperature in degrees Celsius

- salinity = salinity of the water

- phosphate = phosphate concentration in micro moles per liter

- nitrite = nitrite concentration in micro moles per liter

- depth = depth in meters


### The final model with numbers:
`r equatiomatic::extract_eq(final_mdl, wrap = TRUE, use_coefs = TRUE)`
 



