---
title: "Binary Logistic Regression"
description: |
  Palmetto Species Analysis
output: 
  distill::distill_article:
    code_folding: hide
---

## Binary Logistic Regression to Identify Palmetto Species

### Overview
This script uses variables such as plant height, canopy length, canopy width, and the number of green leaves to classify whether palmettos are of species *Serenoa repens* or *Sabal etonia*. Binary logistic regression is used to test feasibility.

Data was collected from 1981 to 2017 in south-central Florida. It includes information regarding year, plant type, species, site, habitat, treatment, survival, height, canopy length, canopy width, number of green leaves, scape, new leaves, canopy, if long, comments, and biomass.

**Data citation:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5

```{r setup, include = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(GGally)
library(broom)
library(jtools)
library(caret)
library(AICcmodavg)
library(data.table)
library(knitr)
```


### Read in the data
```{r}
palmetto <- read_csv(here("data", "palmetto.csv")) # Read in the data
```

### Wrangle the data
Select columns for use in the analysis, change values in the species column to species names, and convert the species column to a factor class.

```{r}
palmetto_clean <- palmetto %>% # Clean the data table
  select(species, height:green_lvs) %>% # Select only the columns I need for analysis, the colon means "all the way over to"
  drop_na()

# Change the values 1 and 2 in the species column to actual species names. 1 = Serenoa repens, 2 = Sabal etonia
palmetto_clean$species <- ifelse(palmetto_clean$species==1,'Serenoa repens','Sabal etonia') 

# Convert the  species column to a factor class
palmetto_clean$species <- as.factor(palmetto_clean$species) 

# levels(palmetto_clean$species) in the console revealed Sabal etonia is factor level 0 and Serenoa repens is factor level 1
```

### Visualize the data to identify trends.

Differences in height, canopy length, canopy width, and green leaves for *Serenoa repens* or *Sabal etonia*.

```{r, include = FALSE}
palmetto_clean %>% 
  ggpairs(aes(color = species)) # Used to get an overview of data before final visualizations are made
```

```{r}
ggplot(data = palmetto_clean, aes(x = height, y = length)) +
  geom_point(aes(color = species)) + 
  labs(title = 'Distribution of Height vs. Canopy Length of Two Palmetto Species', 
       x = 'Plant Height (cm)', 
       y = 'Canopy Length (cm)', 
       color = "Species") + # title, axis labels, and legend title
  theme_minimal() # sets the theme of the graph - for visual purposes
```

**Figure 1.** The relationship between canopy length (cm) and maximum plant height (cm) is shown for each species. *Sabal etonia* is shown in red. *Serenoa repens* is shown in blue.


```{r}
ggplot(data = palmetto_clean, aes(x = width, y = green_lvs)) +
  geom_point(aes(color = species)) + 
  labs(title = 'Distribution of Canopy Width vs. Number of Green Leaves of Two Palmetto Species', 
x = 'Canopy Width (cm)', 
y = 'Number of Green Leaves', 
color = "Species") + # title, axis labels, and legend title
  theme_minimal() # sets the theme of the graph - for visual purposes
```

**Figure 2.** The relationship between the number of green leaves and canopy width (cm) is shown for each species. *Sabal etonia* is shown in red. *Serenoa repens* is shown in blue. 

Based on these plots, the number of green leaves is the most likely variable to help classify the species correctly. It shows the most difference between species. Canopy width, plant height, and canopy length showed similar trends between species.

### Binary logistic regression

```{r}
# Store formulas
f1 <- species ~ height + length + width + green_lvs

f2 <- species ~ height + width + green_lvs
```


```{r}
# Binomial logistic regression for model 1 which includes height + length + width + green_lvs.
palm_blr1 <- glm(formula = f1, 
                    data = palmetto_clean,
                    family = 'binomial')
```


```{r}
# Binomial logistic regression for model 2 which includes height + width + green_lvs and does not include length.
palm_blr2 <- glm(formula = f2, 
                    data = palmetto_clean,
                    family = 'binomial')
```

### Identify which model performs better
Compare the models using AICc

```{r}
AICcmodavg::aictab(list(palm_blr1, palm_blr2)) # this function compares competing models
```

- Model 1 is better because the AIC for model 1 (5194.57) is more than two less than the AIC of model 2 (5987.48).


Compare with a 10-fold cross-validation, using prediction accuracy as our metric.

```{r}
# Using `caret` ("**C**lassification **A**nd **RE**gression **T**raining"):
set.seed(123) 

# tr_ctrl <- trainControl(method = "cv", number = 10)
tr_ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

# Train the model
model1 <- train(f1, data = palmetto_clean, 
               method = "glm", family = 'binomial',
               trControl = tr_ctrl)

model2 <- train(f2, data = palmetto_clean, 
               method = "glm", family = 'binomial',
               trControl = tr_ctrl)
```

- Model 1 (92% accurate) is more accurate than model 2 (90% accurate) according to 10-fold cross validation.

- **Based on both AIC values and 10-fold cross validation, model 1 is the better model.**

### Train model 1 using the entire dataset, and create a finalized table containing the binary logistic regression model results 

```{r}
final_mdl <- glm(formula = f1, 
                 data = palmetto_clean, 
                 family = 'binomial')
```


**Table 1.** Model 1 binary logistic regression model results.

```{r}
### Get a tidy version w/ broom:
final_mdl_tidy <- broom::tidy(final_mdl)

setnames(final_mdl_tidy, old=c("term","estimate", "std.error", "statistic", "p.value"), new=c("Term","Estimate", "Standard Error", "z-value", "p-value"))

kable(final_mdl_tidy)
```

- The model is predicting the likelihood that the species is a *Serenoa repens* because the factor level of *Serenoa repens* is 1.

- All variables are statistically significant.

- Higher length and width would be less likely to be *Serenoa repens* because the value is negative.

- Higher height and more green leaves would be more likely to be a *Serenoa repens* because the value is positive. This makes sense when looking at graphs above.

### Evaluation of how successfully this model would “classify” a plant as the correct species, using a 50% cutoff.

**Table 2.** Evaluation of how successfully this model would “classify” a plant as the correct species, using a 50% cutoff.

```{r}
blr1_fitted <- palm_blr1 %>% # fitted column reports the probability that an individual is a Serenoa repens
  broom::augment(type.predict = "response") %>% # instead of taking the log odds we are taking the odds and converting them into a probability
  mutate(predicted_species = ifelse(.fitted >= 0.5, "Serenoa repens", "Sabal etonia")) %>% 
  group_by(species) %>% # separate by species
  summarize(number_correctly_identified = sum(species == predicted_species),
            number_total = n()) %>% # number_total is needed to identify percent correctly identified
  mutate(percent_correctly_identified = number_correctly_identified/number_total * 100) %>% 
  select(species, number_correctly_identified, percent_correctly_identified)

setnames(blr1_fitted, old=c("species","number_correctly_identified", "percent_correctly_identified"), new=c("Species","Number Correctly Identified", "Percent Correctly Identified"))
kable(blr1_fitted)
```

A binary logistic regression model using plant height, canopy length, canopy width and green leaves as predictor variables is better at predicting palmetto species (*Serenoa repens* and *Sabal etonia*) than a model using plant height, canopy width and the number of green leaves. The model correctly identified 5701 (92.62%) of *Sabal etonia* correctly and 5548 (90.77%) of *Serenoa repens* correctly.
 
 